Think deep for reef resilience

### 2025-08-15 paper/paper_draft.qmd
Initial go at incorporating rf analysis into results section
Have added text about results gained from rf analysis and pdps. Also moved rf figure to supplementary section.

### Partial Dependence Analysis

A Partial Dependence Analysis was conducted to understand the relationship between
individual features and time series cluster predictions using a trained Random Forest (RF)
classifier [@friedman2001]. The analysis utilised the reef characteristics of interest
including the median reef depth, mean DHW for the given scenario, total connectivity
strength, and the outgoing-to-incoming connectivity ratio (\$O:I\$). The full dataset
consisted of the timeseries classification of each of the 2429 reefs under the five GCMs
explored, for a total of 12,145 samples. The dataset was randomly split into training (60%)
and testing (40%) subsets to ensure model generalisability. The RF classifier achieved a 99%
accuracy on the training set and 66% on the test set.

While the 99% accuracy suggests the classifier is overfit to the training data, the 66% test
accuracy indicates a meaningful relationship between modelled reef characteristics and their
assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving
at a similar 67% accuracy rate. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier is reliable.
Furthermore, F1 scores for Low, Medium, and High were determined to be 0.7, 0.57 and 0.68
respectively, indicating balanced performance across the cluster categories. Lower
performance with the Medium cluster reflect the inherent difficulty of classifying
transitional states in natural systems. For management applications, this level of
performance provides valuable decision support when combined with expert knowledge and the
model's probabilistic outputs.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual reef characteristic were systematically (one-at-a-time) varied across
a grid of 50 equally spaced points spanning its range, while all other features were left at
their modelled values. The procedure creates (1) a modified test dataset, where the target
reef characteristic was set to the grid value for all samples, (2) generates probabilistic
predictions for all three cluster classes using the trained RF classifier, and (3) the mean
predicted probability for each class was calculated across all test observations. The median
was also assessed (not shown) but was found not to affect the interpretation of the analysis
results. The process isolates the marginal effect of each feature by averaging over the
natural distribution of all other features in the test dataset.

@@

```{julia}
#| echo: false
#| warning: false
#| message: false
#| results: asis
#| label: tbl-access-esm1-5-bioregion-regression
#| tbl-cap: Multinomial logistic regression coefficients for the statistical model containing median reef depth, base-10-log total connectivity strength, base-10-log source-to-sink ratio and ACCESS-ESM1-5 mean timeseries DHW value as predictor variables. ACCESS-ESM1-5 bioregion scale reef cluster assignment is the response variable with three levels, 'low' (regression reference level), 'medium' and 'high'.

# using PrettyTables, Markdown
# cd("../src")

# include("../src/common.jl")

# cluster_regression = CSV.read(joinpath(output_path, "cluster_regression_outputs/ACCESS-ESM1-5_bioregion_cluster_regression_coefficients.csv"), DataFrame)
# cluster_regression.Name = replace.(cluster_regression.Name, ["ACCESS-ESM1-5_bioregion_cluster_cats: " => "Cluster "])

# for name in names(cluster_regression)
#     if isa(cluster_regression[:, name], Vector{Float64})
#         cluster_regression[:, name] = round.(cluster_regression[:, name]; sigdigits=2)
#     end
# end

# io = IOBuffer()

# header = ["Response level ~ Term", "β estimate", "SE", "t-value", "Pr > |t|", "2.5%", "97.5%"]
# pretty_table(io, cluster_regression, backend=Val(:markdown), header=header)

# md_string = (String(take!(io)))
# Markdown.parse(md_string)
```

\[**BG: Need to add quantification results to text for regressions.**\]

@@

## Partial Dependence

The bioregion variable was excluded from partial dependence analysis due to its categorical
nature providing limited interpretable information. The decision to exclude bioregions from
the analysis is also justified by a low feature importance score.

![Feature importance scores (Panel A) of the Random Forest classifier, and Partial
Dependence Plots for reef characteristics of interest. Results for bioregions were excluded
as they were found to be
uninformative.](../figs/ts_cluster_rf_feature_analysis.png){#ts-cluster-pdp
alt="Feature importance scores (Panel A) of the Random Forest classifier, and Partial Dependence Plots for reef characteristics of interest. Results for bioregions were excluded as they were found to be uninformative."
fig-align="center" width="9.2cm"}

@@

>>>>

#### Partial Dependence Analysis

Partial Dependence Analysis was conducted to understand the relationship between individual
features and time series cluster predictions using a trained Random Forest (RF) classifier
[@friedman2001]. The analysis utilised the reef characteristics of interest including the
median reef depth, mean DHW for the given GCM, total connectivity strength, and the
outgoing-to-incoming connectivity ratio (\$O:I\$). The full dataset consisted of the
timeseries cluster assignment of each of the 2429 reefs under the five GCMs explored, for a
total of 12,145 samples. The dataset was randomly split into training (60%) and testing
(40%) subsets to ensure model generalisability. The RF classifier achieved a 99% accuracy on
the training set and 66% on the test set.

While the 99% accuracy suggests the classifier is overfit to the training data, the 66% test
accuracy indicates a meaningful relationship between modelled reef characteristics and their
assigned clusters has been captured. A 3-fold cross-validation was also conducted, arriving
at a similar 67% accuracy rate. The consistent level of accuracy across test and
cross-validation demonstrates the generalised performance of the RF classifier is reliable.
Furthermore, F1 performance scores for Low, Medium, and High were determined to be 0.7, 0.57
and 0.68 respectively, indicating balanced performance across the cluster categories. Lower
performance with the Medium cluster reflect the inherent difficulty of classifying
transitional states in natural systems. For management applications, this level of
performance provides valuable decision support when combined with expert knowledge and the
model's probabilistic outputs.

Partial dependence analysis was performed by applying the trained model to the test dataset.
Values for individual reef characteristic were systematically (one-at-a-time) varied across
a grid of 50 equally spaced points spanning its range, while all other features were left at
their modelled values. The procedure creates (1) a modified test dataset, where the target
reef characteristic was set to the grid value for all samples, (2) generates probabilistic
predictions for all three cluster classes using the trained RF classifier, and (3) the mean
predicted probability for each class was calculated across all test observations. The median
probability was also assessed (not shown) but was found not to affect the interpretation of
the analysis results. The process isolates the marginal effect of each feature by averaging
over the natural distribution of all other features in the test dataset.

@@

@@

### Partial Dependence

The bioregion variable was excluded from partial dependence analysis due to its categorical
nature providing limited interpretable information. The decision to exclude bioregions from
the analysis is also justified by a low feature importance score (9.9%). Reef median depth
and mean DHW level have the highest feature importance scores, accounting for 25.7% and
23.7% predictive importance respectively. The probability of reefs being clustered into low
and high clusters is greatest at shallow and deep depths respectively, with medium cluster
reefs experiencing lower variation in assignment probability across depth
[@sfig-ts-cluster-pdp]. The carrying capacity (total habitable area) accounts for 14.3%
feature importance. Total strength and source-to-sink ratio connectivity metrics account for
13.3% and 13.0% feature importance respectively. Partial dependence plots for carrying
capacity, total connectivity strength and source-to-sink ratio indicate that their influence
on cluster assignments is concentrated at the low range of predictor values
[@sfig-ts-cluster-pdp]. This aligns with the right-skewed distributions of values across
reefs for these three predictor variables. **\[BG: Will add either predictor distributions
to these sub plots, or just in a general variable distribution plot in the supplementary
material.\]**

@@

![Feature importance scores (Panel A) of the Random Forest classifier, and Partial
Dependence Plots for reef characteristics of interest. Results for bioregions were excluded
as they were found to be
uninformative.](../figs/ts_cluster_rf_feature_analysis.png){#sfig-ts-cluster-pdp
fig-align="center" width="9.2cm"}
